The role of lithic bipolar technology in Western Iberia’s Upper Paleolithic: the case of Vale Boi (southern Portugal)

### 2018-10-07 analysis/paper/paper.Rmd
submitted

---
title: "The role of lithic bipolar technology in Western Iberia’s Upper Paleolithic: the case of Vale Boi (southern Portugal)"
author:
  - Pedro Horta (ICArEHB, University of Algarve, Faro, Portugal) pedroehorta@gmail.com
  - João Cascalheira (ICArEHB, University of Algarve, Faro, Portugal) jmcascalheira@ualg.pt
  - Nuno Bicho (ICArEHB, University of Algarve, Faro, Portugal) nbicho@ualg.pt
date: '`r Sys.Date()`'
output: 
    bookdown::word_document2: #subsistute by pdf_document2 or hmtl_document2 as necessary
      fig_caption: yes
      reference_docx: "../templates/template2.docx" 
bibliography: ScaledPiecesVB.bib
csl: "../templates/springer-humanities-author-date.csl"
abstract: |
  Scaled or splintered pieces are one of the most common lithic artifacts type in Upper Paleolithic assemblages throughout Europe, especially in its westernmost regions. Despite this, even after one century of being identified there is still no consensus on how to define, analyze or even interpret these tools. In western Iberia there is a clear lack of comprehensive studies regarding this type of artifacts at a regional scale. In this paper we present a first techno-morphological analysis of a sample of scaled pieces from the Upper Paleolithic site of Vale Boi. Our first aim was to build upon existing analytical models in order to identify function and possible reduction strategies for these artifacts. Our second goal was to critically evaluate the role of these artifacts within western Iberia’s Upper Paleolithic.
  Our results showed that functional identification of scaled pieces is still not clear. By comparing our data with other author’s we found that current models could not be applied to the archaeological record, as the attribute variability is too high. Furthermore, in this region we found that higher frequencies of bipolar technology can be found related to residential sites due to both functional and cultural patterns. While we still cannot define a specific function for these artifacts (intermediate pieces or wedges for working hard raw materials or cores for the extraction of chips and small bladelets), it is clear that they had a major role in the variability and intensification of resource exploitation during the Upper Paleolithic in western Iberia.

---

@@

library(knitr)
library(tidyverse)
library(broom)
library(FactoMineR)
library(factoextra)
library(ggpubr)
library(ggplot2)
library(ggrepel)
library(maptools)
library(ggmap)
library(devtools)
library(legendMap)
library(tab)
library(sjstats)
library(powerAnalysis)
library(epitools)
#library(ScaledPiecesVB)

@@

```{r VBLocation, echo = FALSE, fig.cap="Location of the site of Vale Boi"}

@@

knitr::include_graphics("../figures/VBMap.png")

@@

All analyses and data processing were accomplished in R (version `r getRversion()`) [@r_core_team_r:_2013]. Following recent concerns on the reproducibility of archaeological analysis we include the entire R code used for all the analysis and visualizations contained in this paper in our supplemental online material (SOM) at https://dx.doi.org/… To produce those files we followed the procedures described by Marwick et al. [-@marwick_packaging_2017] for the creation of research compendiums to enhance the reproducibility of research. The files provided contain all the raw data used in our analysis as well as a custom R package [@wickham_r_2015] holding the code use for all analysis and to produce all tables and figures. To enable maximum re-use, our code is released under the MIT license, our data as CC-0, and our figures as CC-BY, [for more information see @marwick_computational_2016].

@@

assemb <- cross.tb(techno_data, "RawMaterial", var_list)

knitr::kable(assemb, caption= "Frequencies of scaled pieces by raw material and chronology")

@@

knitr::kable(Gravettian, caption= "Technological attribute frequencies by raw materials for the Gravettian sample. Percentages are shown in parentheses.")

@@

knitr::kable(`Proto-Solutrean`, caption= "Technological attribute frequencies by raw materials for the Proto-Solutrean sample. Percentages are shown in parentheses.")

@@

knitr::kable(Solutrean, caption= "Technological attribute frequencies by raw materials for the Solutrean sample. Percentages are shown in parentheses.")

@@

knitr::kable(Magdalenian, caption= "Technological attribute frequencies by raw materials for the Magdalenian sample. Percentages are shown in parentheses.")

@@

```{r meanplots, echo = FALSE, fig.cap="Barplots of means for Area (Length x Width) and Thickness of scaled pieces by raw material and across the four chronological phases. Error bars represent standard deviations."}

@@

```{r plotbyplatforms, echo = FALSE, fig.cap="Boxplot of Area and Thickness distribution for each raw material."}

@@

```{r morphoattributes, echo = FALSE, fig.cap="Frequency of morphological attributes for each raw material. Damaged platforms were grouped so that each artifact was only counted once and to avoid wrong comparisons between active a hammered platforms. A - Distribution of damage; B - Arrangement of scars; C - Extension of scars; D - Delineation of damaged platforms; E - Facial distribution of scars; F - Angle of damaged platforms."}

@@

knitr::kable(scaledpiecesportugal, caption= "Frequencies of scaled pieces in portuguese Upper Paleolithic sites")

@@

Higher degrees of representation of scaled pieces in sites can be originated by factors such as site function, but also by cultural patterns. Regarding site functionality, if we look at the *loci* with higher counts of scaled pieces (Vale Boi’s Terrace and Slope and Fonte Santa), both are open-air residential sites. One example outside of this region is the Southwestern German Magdalenian site of Gönnersdorf, where according to Sano (2012) scaled pieces were the second largest assemblage from the site’s retouched utensils (257 from a total of 1501 utensils), corresponding to c. 17,12% of the total retouched tools assemblage. This typological index is as high as both Vale Boi’s and Fonte Santa’s and like both of these sites, Gönnersdorf is an open-air residential site. However, these data can be misleading, as Casal do Felipe (5,91%), Terra do Manuel 1940-42 (1,03%), Terra do Manuel 1988-89 (2,13%) and Vale Comprido – Encosta (0,9%) are also open-air multifunctional sites. These sites show the same functionality yet, much smaller representations of scaled pieces.

@@

# Acknowledgements

>>>>

---
title: "The role of lithic bipolar technology in Western Iberia’s Upper Paleolithic: the case of Vale Boi (southern Portugal)"
author:
  - Pedro Horta* (ICArEHB, University of Algarve, Faro, Portugal)
  - João Cascalheira (ICArEHB, University of Algarve, Faro, Portugal)
  - Nuno Bicho (ICArEHB, University of Algarve, Faro, Portugal)
  - "*corresponding author | email: pedroehorta@gmail.com | phone: +351 911 005 506"

output: 
    bookdown::word_document2: #subsistute by pdf_document2 or hmtl_document2 as necessary
      fig_caption: yes
      reference_docx: "../templates/template2.docx" 
bibliography: ScaledPiecesVB.bib
csl: "../templates/springer-socpsych-author-date.csl"
---

# Abstract

  Scaled or splintered pieces are one of the most common lithic artifacts type in Upper Paleolithic assemblages throughout Europe, especially in its westernmost regions. Despite this, and even after one century of being identified there is still no consensus on how to define, analyze or interpret these tools. In western Iberia there is a clear lack of comprehensive studies regarding this type of artifacts at a regional scale. In this paper we present a first techno-morphological analysis of a sample of scaled pieces from the Upper Paleolithic site of Vale Boi. Our first aim was to build upon existing analytical models in order to identify function and possible reduction strategies for these artifacts. Our second goal was to critically evaluate the role of these artifacts within western Iberia’s Upper Paleolithic.
  Our results showed that functional identification of scaled pieces is still not clear. By comparing our data with other author’s we found that current models could not be applied to the archaeological record, as the attribute variability is too high. Furthermore, in this region we found that higher frequencies of bipolar technology can be found related to residential sites due to both functional and cultural patterns. While we still cannot define a specific function for these artifacts (intermediate pieces or wedges for working hard raw materials or cores for the extraction of chips and small bladelets), it is clear that they had a major role in the variability and intensification of resource exploitation during the Upper Paleolithic in western Iberia.

# Keywords

Stone tools; Bipolar technology; Upper Paleolithic; Western Europe

@@

#Please see the functions.R file in /R directory for other libraries used for this compendium

library(knitr)
library(citr)
library(tidyverse)
library(broom)
library(FactoMineR)
library(factoextra)
library(ggpubr)
library(ggplot2)
library(ggrepel)
library(maptools)
library(ggmap)
library(devtools)
library(legendMap)
library(tab)
library(sjstats)
library(powerAnalysis)
library(epitools)
library(wesanderson)

#library(ScaledPiecesVB)

@@

```{r VBLocation, echo = FALSE, fig.cap="Location of the site of Vale Boi. Map data are from stamenmap, using the ggmap package Kahle and Wickham [-@kahle_ggmap:_2013]"}

@@

knitr::include_graphics("../figures/location_map.png")

@@

All analyses and data processing were accomplished in R (version `r getRversion()`) [@r_core_team_r:_2013]. Following recent concerns on the reproducibility of archaeological analysis we include the entire R code used for all the analysis and visualizations contained in this paper in our supplemental online material (SOM) at https://dx.doi.org/10.17605/OSF.IO/WPXGH To produce those files we followed the procedures described by Marwick et al. [-@marwick_packaging_2017] for the creation of research compendiums to enhance the reproducibility of research. The files provided contain all the raw data used in our analysis as well as a custom R package [@wickham_r_2015] holding the code use for all analysis and to produce all tables and figures. To enable maximum re-use, our code is released under the MIT license, our data as CC-0, and our figures as CC-BY, [for more information see @marwick_computational_2016].

@@

assemb <- as.tibble(cross.tb(techno_data, "RawMaterial", var_list))

assemb <- assemb %>%
  select(" " = "Variable", "Chert", "Quartz", "Chalcedony", "Total" = "Overall") %>% 
  slice(2:5)

assemb <- assemb[c(1,3,4,2),]

knitr::kable(assemb, caption= "Frequencies of scaled pieces used in this study, by raw material and chronological period. Percentages are shown in parentheses")

@@

knitr::kable(Gravettian, caption= "Technological attributes frequencies by raw materials for the Gravettian sample. Percentages are shown in parentheses.")

@@

knitr::kable(`Proto-Solutrean`, caption= "Technological attributes frequencies by raw materials for the Proto-Solutrean sample. Percentages are shown in parentheses.")

@@

knitr::kable(Solutrean, caption= "Technological attributes frequencies by raw materials for the Solutrean sample. Percentages are shown in parentheses.")

@@

knitr::kable(Magdalenian, caption= "Technological attributes frequencies by raw materials for the Magdalenian sample. Percentages are shown in parentheses.")

@@

```{r meanplots, echo = FALSE, fig.cap="Barplots of means for Area (Length x Width) and Thickness of scaled pieces, by raw material and across the four chronological phases. Error bars represent standard deviations."}

@@

```{r plotbyplatforms, echo = FALSE, fig.cap="Boxplot of Area (Length x Width) and Thickness distribution for each raw material."}

@@

```{r morphoattributes, echo = FALSE, fig.cap="Frequency of morphological attributes for each raw material. Opposed damaged platforms were grouped so that each artifact was only counted once and to avoid wrong comparisons between active a hammered platforms. A - Distribution of damage; B - Arrangement of scars; C - Extension of scars; D - Delineation of damaged platforms; E - Facial distribution of scars; F - Angle of damaged platforms."}

@@

knitr::kable(scaledpiecesportugal, caption= "Frequencies of scaled pieces and bipolar cores in portuguese Upper Paleolithic sites.")

@@

Higher degrees of representation of scaled pieces in sites can be originated by factors such as site function, but also by cultural patterns. Regarding site functionality, if we look at the *loci* with higher counts of scaled pieces (Vale Boi’s Terrace and Slope and Fonte Santa), all are open-air residential sites. One example outside of this region is the Magdalenian site of Gönnersdorf in Southwestern Germany, where according to Sano (2012) scaled pieces were the second most frequent tool types within the site’s retouched tools assemblage (257 from a total of 1501 utensils, corresponding to c. 17,12%). This frequency is as high as both Vale Boi’s and Fonte Santa’s, and like both of these sites Gönnersdorf is an open-air residential site. However, these data can be misleading, as Casal do Felipe (5,91%), Terra do Manuel 1940-42 (1,03%), Terra do Manuel 1988-89 (2,13%) and Vale Comprido – Encosta (0,9%) are also open-air multifunctional sites with much smaller representations of scaled pieces.

@@

# Acknowledgements
We would like to thank a diversity of institutions that provided funding to the work in Vale Boi in the last decade. These are Archaeological Association of Algarve, Fundaç~ao para a Ciência e Tecnologia, Instituto Português de Arqueologia, National Geographic Society, and Wenner-Gren Foundation for Anthropological Research.
